---
output: pdf_document
latex_engine: xelatex
fontsize: 11pt
linestretch: 1.5
fig_width: 10
fig_height: 6
header-includes:
- \usepackage{amssymb}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{enumitem}
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{multicol}
- \setlength{\columnsep}{-20pt}
- \usepackage{titling}
- \usepackage{color}
- \usepackage{tikz}
- \usetikzlibrary{trees}
- \usepackage{graphicx}
- \usepackage{lscape}
- \usepackage{longtable}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---



\begin{center}
\thispagestyle{empty}
\vspace{10pt}
\normalsize
\textbf{Illiteracy and indigenous education:} \\
\textbf{Impact evaluation of subsidies for indigenous schools in Mexico} \\
\vspace{20pt}
\textbf{Grimaldi-Ferreira, Alejandro$^{a}$} \\
\vspace{20pt}
\end{center}

 $^{a}$ Tecnología e Información para la Salud, S.A. de C.V., Zip Code 06170, Mexico. ale_grimaldi@hotmail.com


# Abstract

In Mexico, 27.3% of indigenous individuals aged 15 and over were illiterate in 2010, compared to a rate of 6.9% in the general population. Between 2014 and 2021, a federal program called the Educational Inclusion and Equity Program provided subsidies to elementary schools for indigenous education. I evaluate the impact of the program using a difference-in-difference identification strategy and census and administrative data. I find that the program increased literacy among the population aged 6 to 11 in each municipality by 3.42 percentage points from 2014 to 2020, while also increasing school attendance, enrollment, and the number of classrooms in indigenous schools.


# Keywords

Educational economics, human capital, state and federal aid.


# Subject Descriptors

National Government Expenditures and Education H52

Economic Development: Human Resources; Human Development; Income Distribution; Migration O15


# Abbreviations

CONEVAL: National Council for the Evaluation of Social Development Policy

INEGI: National Institute of Statistics and Geography

PIEE: Educational Inclusion and Equity Program

SEP: Ministry of Public Education


# Acknowledgements

This research did not receive any specific grant from funding agencies in the public, commercial, or not-for-profit sectors. I thank Arturo Aguilar, Mauricio Romero, Claudia Aburto, Manett Vargas, and René Livas for helpful comments. All errors are my own.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
library(pacman)
#install.packages("remotes")
#install.packages("withr")
#remotes::install_github("diegovalle/mxmaps", force = T)
#library(devtools)
#install_github("GeoBosh/Rdpack")
p_load(rmarkdown, haven, plm, car, stargazer, dplyr, DataCombine, clubSandwich, ggplot2, stats, knitr, kableExtra, mxmaps, taRifx, fixest)
```


```{r echo=FALSE}
input <- "C:/Users/ale_g/OneDrive/Documentos/Economics of Education Review/Bases de Datos"

datos <- read.csv(paste(input, "Analfabetismo y Educación Indígena 2010-2020.csv", sep="/"))
datos <- as.data.frame(datos)
datos$X <- NULL

grupos <- read.csv(paste(input, "Analfabetismo y Educación Indígena 2000-2020 (Grupos).csv", sep="/"))
grupos <- as.data.frame(grupos)
grupos$X <- NULL

edades <- read.csv(paste(input, "Analfabetismo y Educación Indígena 2000-2020 (Edades).csv", sep="/"))
edades <- as.data.frame(edades)
edades$X <- NULL

mujeres <- read.csv(paste(input, "Analfabetismo y Educación Indígena 2000-2020 (Mujeres).csv", sep="/"))
mujeres <- as.data.frame(mujeres)
mujeres$X <- NULL

f911 <- read.csv(paste(input, "Analfabetismo y Educación Indígena 2000-2020 (Formato 911).csv", sep="/"))
f911 <- as.data.frame(f911)
f911$X <- NULL
```


\newpage
\setcounter{page}{1}


# 1. Introduction

Literacy is a cornerstone for human and social capital. Illiteracy results not only in the deprivation of the right to education, but it is also an important channel for poverty. Lacking access to education generally, and access to literacy specifically, has implications for the person's economic well-being and their access to other social rights.

> *"Literacy is a fundamental skill, which provides the foundation for further learning and enables students to access the printed texts essential for further education, for participation in most jobs, and for informed political participation".* (Reimers 2006, p. 196).

There is extensive literature on the impact of teaching in a language known by students. It impacts students' academic achievement (Bühmann & Trudell 2007; Miscevic-Kadijevic 2015) and school attendance (Smits, Huisman & Kruijff 2008).

Furthermore, in studies examining primary instruction in native language vis-á-vis Spanish, the evidence suggests that basic education should be based on students' native language (Bühmann & Trudell 2007; Ganuza & Hedman 2017; Nkonde, et al. 2018), while Spanish is learned as a second language during the first years. If, eventually, middle or high school is reached without having learned Spanish, school performance will be lower on average than that of bilingual students or those whose native language is Spanish.

In the case of Mexico, the indigenous population is disproportionately affected by this problem. In 2010, the percentage of the population aged 15 or over with illiteracy stood at 6.9%. However, among the population that speaks an indigenous language, this rate rises to 27.3% (Narro & Moctezuma, 2012).

Between 2014 and 2021, a federal program called Educational Inclusion and Equity Program (PIEE for its acronym in Spanish), coordinated by the Ministry of Public Education (SEP for its acronym in Spanish), provided subsidies to elementary schools with instruction in indigenous languages. Subsidies were allocated to any indigenous school requesting them. This paper evaluates the impact of these federal subsidies using information from the 2000, 2005, 2010, 2015 and 2020 Population Censuses, Counts and Intercensal Surveys, as well as on the municipal coverage of the program between 2014 and 2019.

To identify the impact of this policy, I use a differences-in-differences strategy. The presence of the program reduced the percentage of illiteracy among the population aged 6 to 11 in municipalities with indigenous education schools by 3.42 percentage points from 2010 to 2020. The causal mechanism is likely due to an increase in school infrastructure. I find positive impacts using several proxy variables were used as outcomes, like school attendance (1.31 pp), school enrollment (6.78 students), number of classrooms (0.54), and number of teachers. For each specification, placebo effects between 2000 and 2010 were incorporated in order to provide evidence on the parallel trends assumption. In general, the assumption seems plausible in light of the fact that placebo coefficients are small and lack statistical significance.

The rest of the paper is organized as follows. Section 2 provides a theoretical framework. Section 3 describes the data and performs summary statistics. Section 4 describes the empirical strategy. Section 5 presents the results. Section 6 concludes.


\newpage


# 2. Theoretical Framework

In Mexico, there was a social program focused on indigenous elementary education: the PIEE coordinated by the SEP. This program emerged in 2014 as a merger of seven programs and had six components: i) support for indigenous education schools, ii) support for educational centers for migrants, iii) support for special education services, iv) support for *telesecundaria* education, v) support for Higher Secondary Education Centers for Students with Disabilities and vi) support for Public Institutions of Higher Education for the inclusion of students with disabilities or in contexts of vulnerability.

This study focuses solely on the support component for indigenous education schools, under the responsibility of the General Directorate of Indigenous Education of the Undersecretary of Basic Education. This program intended to address as "inadequate educational offer to the diverse needs of children and young in a vulnerable context" (SEP 2014; p. 21). Thus, the target population was defined as "all Basic Education schools that offer education services to indigenous people, migrants, special education and *telesecundarias* that decide to participate in the program" (SEP 2014, p. 25).

Regarding the coverage of the program, Figure 1 shows that it had a high concentration over the states of the southeast, as well as some municipalities of Puebla, Oaxaca and other states through which the Sierra Madre Occidental is located. According to Figure 2, this map is quite similar to the distribution of the population that speaks an indigenous language in Mexico, which is not surprising, since the subsidies are intended precisely to support indigenous education schools. Between 2014 and 2019, 33.36% of the municipalities received subsidies from the program (817 municipalities).


```{r results="asis", echo=FALSE, warning=FALSE}
temp <- subset(grupos, anio == 2015 & grupo == "edad_6a11")
temp <- temp %>% select(mun, trat)
temp <- rename(temp, region = mun)
temp <- rename(temp, value = trat)

mxmunicipio_choropleth(temp,
                       title = "Figure 1. Coverage of the program between 2014 and 2019")

temp <- subset(grupos, anio == 2010 & grupo == "edad_6a11")
temp <- temp %>% select(mun, hlengua)
temp <- rename(temp, region = mun)
temp <- rename(temp, value = hlengua)

mxmunicipio_choropleth(temp, 
                       num_colors = 1,
                       title = "Figure 2. Percentage of children aged 6 to 11 who speaks an indigenous\nlanguage by municipality in 2010",
                       legend = "%")
```


The program delivered subsidies to elementary schools of indigenous education that request it (through the Local Educational Authorities), which could have as destination (DOF 2019):

\begin{enumerate}
\item Actions for academic strengthening: diplomas, congresses, seminars, travel expenses and transportation, subsidies for continuing education for indigenous teachers, advice, and technical assistance.
\item Contextualization actions: adaptation of content and materials, design of teaching methods and strategies, implementation of reading and writing programs in indigenous language, and purchase of educational materials.
\end{enumerate}

Regarding the "contextualization actions", Duarte, Moreno, and Gargiulo (2011) argue that the access to electricity, telephony, drinking water, sewage and toilets services are decisive for the achievement of students in language and mathematics tests. Furthermore, Barrett, et al. (2019) point out that installing protection against natural disasters, access to basic services (electricity, water, sanitation, and communications), spaces for outdoor recreation and good and safe physical conditions in schools are elements that have a positive impact on school attendance of students and teachers, student health, and the likelihood that teachers will remain in the profession.

> *"Investments in school infrastructure and the physical conditions for learning are not luxury but a need".* (Barrett, et al. 2019, p. 13).

Not only the infrastructure of schools matters, but also the acquisition of educational material, such as textbooks. Evans, et al. (2010) show that children who grow up in homes with many books acquire, on average, three more years of schooling than children in homes without books.

However, Glewwe, Kremer, and Moulin (2009) argue that students with poorer academic backgrounds do not benefit from purchasing textbooks in schools. In fact, the underlying program that randomly handed out textbooks in Kenya had no impact on learning in the aggregate.

Likewise, regarding *input-based* actions, De Hoyos, Holland, and Troiano (2015) show that policies based on increasing the number of teachers per group and their salary, improving school infrastructure, and implementing the use of computers did not lead to better results in standardized tests in Argentina.

> *"This is not to say that education inputs are not important in the learning process, but their impact is limited and seems to diminish over time because it is subject to the systems ability to use them in an effective way".* (De Hoyos, Holland, and Troiano 2015).

That is, policies based on the acquisition and improvement of educational infrastructure depend on the institutional and teaching capacity to transform it into student learning. This is strongly linked to the relevance that the PIEE's "actions for academic strengthening" would have.

The "actions for academic strengthening" of the program could be used to subsidize the continuous training of teachers. Improving teaching quality can have a significant impact on learning (Rivkin, Hanushek, and Kain 2005; Bruns & Luque 2015), because they are the main input in the educational process. Hanushek, Piopiunik, and Wiederhold (2014) show that teachers' cognitive abilities are highly correlated with student outcomes.

In summary, the program subsidies could well have had a positive impact on the learning of students in indigenous schools. However, it is also likely that these subsidies are, by themselves, insufficient, especially in contexts where there were teachers with a lower pedagogical level or a wealth of human capital and the "actions for academic strengthening" would not have been sufficient to improve continuous teacher training.


\newpage


# 3. Data and Descriptive Analysis

To carry out the empirical analysis, I use data from the Population Censuses, Counts and Intercensal Surveys of 2000, 2005, 2010, 2015, and 2020. I analyze data at the municipal level on a total of 2,419 municipalities.

I compile variables on the literacy rates, population that speaks an indigenous language, female share of the population, school attendance, share of rural localities, and average age and years of schooling. A locality is defined as rural if its population is less than 2,500 people.

To incorporate information for other control variables, I use data from the Multidimensional Measurement of Poverty of the National Council for the Evaluation of Social Development Policy (CONEVAL). I obtain the percentages of poverty and extreme poverty by municipality, as well as percentages of population with income below the welfare line, population lacking access to social rights (education, health, housing, basic housing services and food), population counts, and Gini index at the municipal level for 2010, 2015 and 2020.

To carry out the empirical analysis, I merge information on the municipalities that received resources from the indigenous education component of the PIEE from 2014 to 2019. This data was obtained through application number 0001100442220 and review resource RRA11287/20, through the National Transparency Platform of the National Institute of Transparency, Access to Information and Protection of Personal Data (INAI).

Finally, to test the causal mechanism, I use multiple proxies for school infrastructure leveraging municipal panel data. I construct the percentage of school attendance by municipality (of children between 6 and 11 years old) using the INEGI Counts, Censuses and Intercensal Survey. I also leverage yearly data from SEP Format 911, exclusively among indigenous education schools, to measure school enrollment, number of teachers, and number of classrooms per school in treated and untreated municipalities from 2000 to 2019.

In summary, the following 4 sources of information were used for the data:
\begin{enumerate}
\item Population Censuses, Counts and Intercensal Surveys of 2000, 2005, 2010, 2015, and 2020 (INEGI).
\item Measurement of poverty and Gini index for 2010, 2015 and 2020 (CONEVAL).
\item Coverage of the program from 2014 to 2019 (INEGI).
\item Administrative data as proxies for school infrastructure from 2000 to 2019 (SEP Format 911).
\end{enumerate}

All data I used, as well as the R Markdown code that generated this article, are available in the following GitHub repository: https://github.com/alegrim13/PIEE_Impact_Evaluation.

Figure 3 shows a stark negative correlation between literacy rates and indigenous status in a cross-section of municipalities in 2020. The higher the percentage of the population that speaks an indigenous language, the lower the level of literacy in the municipality.

Also, Table 1 shows the difference in means before PIEE's application between the control and program municipalities. The municipalities in the program have a lower academic level and school attendance and greater poverty, extreme poverty, population count, and social deprivation.

\newpage


```{r results="asis", echo=FALSE}
temp <- subset(grupos, anio == 2010 & grupo == "edad_6a11")

tabla <- temp %>%
  select(alfabet, hlengua, edad, mujer, asisten, nivacad, rural, pobreza, pobreza_ext, pob, rezed, salud, seg_soc, vivienda, serv_viv, alim, inf_lb, gini, trat) %>%
  group_by(trat) %>%
  summarize(alfabet = mean(alfabet),
            hlengua = mean(hlengua),
            edad = mean(edad),
            mujer = mean(mujer),
            asisten = mean(asisten),
            nivacad = mean(nivacad),
            rural = mean(rural),
            pobreza = mean(pobreza),
            pobreza_ext = mean(pobreza_ext),
            pob = mean(pob),
            rezed = mean(rezed),
            salud = mean(salud),
            seg_soc = mean(seg_soc),
            vivienda = mean(vivienda),
            serv_viv = mean(serv_viv),
            alim = mean(alim),
            inf_lb = mean(inf_lb),
            gini = mean(gini)) %>%
  ungroup() %>%
  select(-trat)

names <- c("Literacy", "Indigenous language speakers", "Age", "Women", "School attendance", "Academic level", "Rural", "Poverty", "Extreme poverty", "Population", "Educational lag", "Lack of health services", "Lack of social security", "Lack of housing", "Lack of housing services", "Lack of food", "Income below welfare line", "Gini index")

tabla <- tabla %>%
  t() %>%
  data.frame(names)

colnames(tabla) <- c("Control", "Program", "Variables")

tabla <- tabla %>%
  select(Variables, Control, Program)

tabla <- tabla %>%
  mutate(Difference = Program - Control) %>%
  mutate(P_Value = c(t.test(alfabet ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(hlengua ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(edad ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(mujer ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(asisten ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(nivacad ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(rural ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(pobreza ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(pobreza_ext ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(pob ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(rezed ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(salud ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(seg_soc ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(vivienda ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(serv_viv ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(alim ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(inf_lb ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value,
                     t.test(gini ~ trat, subset(temp, anio==2010), conf.level = 0.95)$p.value))

rownames(tabla) <- NULL

tabla %>%
  kbl(digits=4, caption = "Difference in means between control and program municipalities in 2010 (population aged 6 to 11 years)") %>%
  kable_classic_2(full_width = F)
```


```{r echo=FALSE}
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(subset(datos, anio==2020), aes(x=hlengua, y=alfabet)) +
  geom_point(colour="dodgerblue4") +
  geom_smooth(method="lm", formula=y~x, colour="darkred") +
  ggtitle("Figure 3. Relationship between indigenous language speakers percentage\nand literacy percentage in 2020") +
  xlab("Indigenous language speakers percentage") + ylab("Literacy percentage")
```


\newpage


# 4. Empirical Strategy

Because schools apply for PIEE subsidies, assignment to treatment is not random. Simple mean differences between municipalities that received the program versus those that did not would be overstated because of selection bias. However, if there is an untreated and a treatment group, panel data for both groups before and after the application of the treatment, there are no time-varying differences between the two groups, and all municipalities had access to the program at the same time, it is possible to use a difference-in-differences specification to identify the causal effects of the treatment. That is, I incorporate the interaction between the treatment dummy at the municipal level and the dummy for the time after the program implementation.

The difference-in-differences framework estimates the difference between the mean of the *outcome* of interest, in this case literacy, between the program group and a counterfactual that is not precisely a pure control group, but the hypothetical scenario about what would have happened with the program group having followed the same behavior as the control group. Hence, this analysis assumes that conditional independence is satisfied.

The empirical framework to be estimated is as follows:

\begin{center}
$Y_{i,t} = \beta_0 + \beta_1 P_{i} + \beta_2 Post_{t} + \beta_{DD} (P_{i} * Post_{t}) + \gamma_i + \lambda_t + U_{i,t}$
\end{center}

where $Y_{i,t}$ is the percentage of literacy in municipality i and year t, $P_{i}$ is an indicator for whether municipality i was part of the program, $Post_{t}$ is a dummy that indicates if the year t is after the application of the program (in this case, after 2013), $P_{i} * Post_{t}$ is the interaction between the two previous variables and, therefore, is a dummy that indicates whether municipality i was part of the program and, furthermore, if year t is after its application. $\gamma_i$ and $\lambda_t$ are individual and time fixed effects. $U_{i,t}$ is the error term in municipality i and year t.

In this way, the element of the interaction or double difference ($P_{i} * Post_{t}$) captures the causal effect of the program as an Average Treatment Effect on Treated (ATT):

\begin{center}
$\beta_{DD} = [E(Y_{i,t} | P_i=1, Post_{t}=1) - E(Y_{i,t} | P_i=1, Post_{t}=0)] - [E(Y_{i,t} | P_i=0, Post_{t}=1) - E(Y_{i,t} | P_i=0, Post_{t}=0)]$
\end{center}

To verify the credibility of the parallel trends assumption, I test for placebo effects in periods prior to the application of the program, from 2000 to 2010.


\newpage


# 5. Results

## 5.1. The impact of the Educational Equity and Inclusion Program

Table 2 shows the results for multiple differences-in-differences specifications for literacy rates at the municipal level for children aged 6 to 11. The period before the program corresponds to 2010. Hence, the program had been in operation for about two years by 2015, and for approximately four years by 2020.

The first column displays results from the pooled ordinary least squares specification. The second incorporates fixed effects by municipality and time, the third controls for the percentage of the population that speaks an indigenous language, the fourth controls for the rest of the socio-demographic variables\footnote{These variables come from CONEVAL's data from 2010, 2015 and 2020, however they are still the same 2,419 municipalities.}, and the fifth is a random effects specification.

In all cases, I cluster standard errors at the municipality level. Poverty controls include percentages of poverty and extreme poverty and population with income below the welfare line or lacking access to education, health, housing, basic housing services and food for 2010, 2015 and 2020.

All specifications yield similar average impacts for the PIEE program and are significant at the 1% level. On average, the presence of the program increased literacy in each municipality by 3.424 percentage points from 2010 to 2020. Disaggregating the effect, the increase was greater between 2014 and 2015 (2.90 percentage points) than that of the period 2016 to 2019 (0.53 percentage points). This does not imply that the effect of the program was reduced: in graph 2, it is observed that these 0.53 points are, rather, an increase on top of the initial effect.

To test for parallel trends, I also show estimates for 2000 and 2005. I detect no significant differences for the coefficients of the placebo effect (except for a significant difference in 2010 in the third column, albeit at the 10% level). Furthermore, the difference in difference coefficient does not seem to change much in sign, magnitude, and significance when controls are incorporated. Overall, these results lend credibility to the parallel trends assumption in this context.


```{r include=FALSE}
### Regresiones por grupos.

### Grupo etario: 6 a 11 años.

temp <- subset(grupos, grupo == "edad_6a11")

temp$dummy_00 <- 0
temp$dummy_00[temp$anio == 2000] <- 1

temp$dummy_05 <- 0
temp$dummy_05[temp$anio == 2005] <- 1

temp$dummy_15 <- 0
temp$dummy_15[temp$anio == 2015] <- 1

temp$dummy_20 <- 0
temp$dummy_20[temp$anio == 2020] <- 1

temp$diff_00 <- temp$trat * temp$dummy_00
temp$diff_05 <- temp$trat * temp$dummy_05
temp$diff_15 <- temp$trat * temp$dummy_15
temp$diff_20 <- temp$trat * temp$dummy_20

# 1. Pooled OLS.

modelo1 <- lm(alfabet ~ trat + diff_00 + diff_05 + diff_15 + diff_20 + factor(anio), temp)
cov1 <- vcovCR(modelo1, type="CR1", cluster=temp$mun)
robust_se1 <- sqrt(diag(cov1))

# 2. Efectos Fijos.

p_datos <- pdata.frame(temp, index = c("mun", "anio"))
is.pbalanced(p_datos)
p_datos <- p_datos[order(p_datos$mun, p_datos$anio),]

modelo2 <- plm(alfabet ~ diff_00 + diff_05 + diff_15 + diff_20 + factor(anio), p_datos, model = "within")
cov2 <- vcovCR(modelo2, type="CR1", cluster=p_datos$mun)
robust_se2 <- sqrt(diag(cov2))

# 3. Efectos Fijos controlando Hablantes de lengua indígena.

modelo3 <- plm(alfabet ~ diff_00 + diff_05 + diff_15 + diff_20 + factor(anio) + hlengua, p_datos, model = "within")
cov3 <- vcovCR(modelo3, type="CR1", cluster=p_datos$mun)
robust_se3 <- sqrt(diag(cov3))

# 4. Efectos Fijos con otros controles.

modelo4 <- plm(alfabet ~ diff_15 + diff_20 + factor(anio) + hlengua + mujer + asisten + nivacad + rural + pobreza + pobreza_ext + pob + rezed + salud + seg_soc + vivienda + serv_viv + alim + inf_lb + gini, p_datos, model = "within")
cov4 <- vcovCR(modelo4, type="CR1", cluster=p_datos$mun)
robust_se4 <- sqrt(diag(cov3))

# 5. Efectos Aleatorios.

modelo5 <- plm(alfabet ~ trat + diff_00 + diff_05 + diff_15 + diff_20 + factor(anio), p_datos, model = "random")
cov5 <- vcovCR(modelo5, type="CR1", cluster=p_datos$mun)
robust_se5 <- sqrt(diag(cov5))

# Prueba de Hausman.

phtest(modelo2, modelo5)
```


```{r results="asis", echo=FALSE}
stargazer(modelo1, modelo2, modelo3, modelo4, modelo5,
          title="Differences in Differences over literacy",
          se=list(robust_se1, robust_se2, robust_se3, robust_se4, robust_se5),
          omit=c("trat", "anio", "mujer", "asisten", "nivacad", "rural", "pobreza", "pobreza_ext", "pob", "rezed", "salud", "seg_soc", "vivienda", "serv_viv", "alim", "inf_lb", "gini"),
          omit.stat=c("f", "adj.rsq", "ser"),
          font.size="scriptsize",
          column.sep.width="0pt",
          model.names=F,
          column.labels=c("Pooled OLS", "Fixed Effects", "Fixed Effects", "Fixed Effects", "Random Effects"),
          add.lines=list(c("Time FE", "Yes", "Yes", "Yes", "Yes", "Yes"),
                         c("Individual FE", "No", "Yes", "Yes", "Yes", "No"),
                         c("Poverty and Gini controls", "No", "No", "No", "Yes", "No")),
          covariate.labels = c("Program dummy * 2000", "Program dummy * 2005", "Program dummy * 2015", "Program dummy * 2020", "Indigenous language speakers"),
          dep.var.labels="Literacy (6 to 11 years old)", no.space=TRUE, header=FALSE)
```


To visualize this effect, Figure 4 displays the *event study* to visualize the effect of the program before (placebo) and after its application in 2014 for each year, incorporating 95% confidence intervals for the estimated ATT of each year.
\break


```{r include=FALSE}
### Event Study:

temp <- subset(grupos, grupo == "edad_6a11")

temp$time_to_treat <- temp$anio - 2014

modelo_tt <- feols(alfabet ~ i(time_to_treat, trat, ref = -4) + factor(anio) + factor(mun), cluster = ~ mun, data = temp)
```


```{r results="asis", echo=FALSE, warning=FALSE}
iplot(modelo_tt, xlab = "Time to program", ylab = "ATT estimate (95% intervals)", main = "Figure 4. Event study of the program over literacy\n(6 to 11 years old)")
```


\newpage


## 5.2. Causal Mechanisms

Ideally, the best alternative to test the causal mechanism between program subsidies and literacy would be using data on i) school infrastructure and ii) teacher quality. However, data on school infrastructure is limited: there is a *Census of schools, teachers and students of basic and special education*, but it was only carried out in 2013. On the other hand, there is information about the evaluation of the Professional Teaching Service, but it is only representative for federal entities and not municipalities.

Another alternative to indirectly measure teaching quality is standardized test results for elementary school students. However, the ENLACE test was carried out until 2013 and was replaced by PLANEA for the 2014-2015 school year, while the PIEE program was born in 2014. Their methodologies are not comparable and a jump in 2014 could be due to a methodological change from ENLACE to PLANEA.

Due to these reasons, multiple proxies were used, mainly in terms of school infrastructure for panel municipal data. The percentage of school attendance by municipality (of children between 6 and 11 years old) was constructed using the INEGI Counts, Censuses and Intercensal Survey.

Due to these reasons, I use multiple proxies for school infrastructure leveraging municipal panel data. I construct the percentage of school attendance by municipality (of children between 6 and 11 years old) using the INEGI Counts, Censuses and Intercensal Survey. I also leverage yearly data from SEP Format 911, exclusively among indigenous education schools, to measure school enrollment, number of teachers, and number of classrooms per school in treated and untreated municipalities from 2000 to 2019.

### 5.2.1. School attendance

Figure 5 displays results from a differences-in-differences regression using school attendance between children aged 6 to 11 as the outcome variable. According to these estimates, the program increased school attendance by 1.31 percentage points in municipalities whose indigenous education schools received subsidies. However, as shown in Table 3, placebo coefficients for the year 2000 are statistically significant at the 1% level across all specifications.


```{r include=FALSE}
### Event Study:

temp <- subset(grupos, grupo == "edad_6a11")

temp$time_to_treat <- temp$anio - 2014

modelo_tt <- feols(asisten ~ i(time_to_treat, trat, ref = -4) + factor(anio) + factor(mun), cluster = ~ mun, data = temp)
```


```{r results="asis", echo=FALSE, warning=FALSE}
iplot(modelo_tt, xlab = "Time to program", ylab = "ATT estimate (95% intervals)", main = "Figure 5. Event study of the program over school attendance\n(6 to 11 years old)")
```


```{r include=FALSE}
### Regresiones por grupos.

### Grupo etario: 6 a 11 años.

temp <- subset(grupos, grupo == "edad_6a11")

temp$dummy_00 <- 0
temp$dummy_00[temp$anio == 2000] <- 1

temp$dummy_05 <- 0
temp$dummy_05[temp$anio == 2005] <- 1

temp$dummy_15 <- 0
temp$dummy_15[temp$anio == 2015] <- 1

temp$dummy_20 <- 0
temp$dummy_20[temp$anio == 2020] <- 1

temp$diff_00 <- temp$trat * temp$dummy_00
temp$diff_05 <- temp$trat * temp$dummy_05
temp$diff_15 <- temp$trat * temp$dummy_15
temp$diff_20 <- temp$trat * temp$dummy_20

# 1. Pooled OLS.

modelo1 <- lm(asisten ~ trat + diff_00 + diff_05 + diff_15 + diff_20 + factor(anio), temp)
cov1 <- vcovCR(modelo1, type="CR1", cluster=temp$mun)
robust_se1 <- sqrt(diag(cov1))

# 2. Efectos Fijos.

p_datos <- pdata.frame(temp, index = c("mun", "anio"))
is.pbalanced(p_datos)
p_datos <- p_datos[order(p_datos$mun, p_datos$anio),]

modelo2 <- plm(asisten ~ diff_00 + diff_05 + diff_15 + diff_20 + factor(anio), p_datos, model = "within")
cov2 <- vcovCR(modelo2, type="CR1", cluster=p_datos$mun)
robust_se2 <- sqrt(diag(cov2))

# 3. Efectos Fijos controlando Hablantes de lengua indígena.

modelo3 <- plm(asisten ~ diff_00 + diff_05 + diff_15 + diff_20 + factor(anio) + hlengua, p_datos, model = "within")
cov3 <- vcovCR(modelo3, type="CR1", cluster=p_datos$mun)
robust_se3 <- sqrt(diag(cov3))

# 4. Efectos Fijos con otros controles.

modelo4 <- plm(asisten ~ diff_15 + diff_20 + factor(anio) + hlengua + mujer + nivacad + rural + pobreza + pobreza_ext + pob + rezed + salud + seg_soc + vivienda + serv_viv + alim + inf_lb + gini, p_datos, model = "within")
cov4 <- vcovCR(modelo4, type="CR1", cluster=p_datos$mun)
robust_se4 <- sqrt(diag(cov3))

# 5. Efectos Aleatorios

modelo5 <- plm(asisten ~ trat + diff_00 + diff_05 + diff_15 + diff_20 + factor(anio), p_datos, model = "random")
cov5 <- vcovCR(modelo5, type="CR1", cluster=p_datos$mun)
robust_se5 <- sqrt(diag(cov5))

# Prueba de Hausman.

phtest(modelo2, modelo5)
```


```{r results="asis", echo=FALSE}
stargazer(modelo1, modelo2, modelo3, modelo4, modelo5,
          title="Differences in Differences over school attendance",
          se=list(robust_se1, robust_se2, robust_se3, robust_se4, robust_se5),
          omit=c("trat", "anio", "mujer", "asisten", "nivacad", "rural", "pobreza", "pobreza_ext", "pob", "rezed", "salud", "seg_soc", "vivienda", "serv_viv", "alim", "inf_lb", "gini"),
          omit.stat=c("f", "adj.rsq", "ser"),
          font.size="scriptsize",
          column.sep.width="0pt",
          model.names=F,
          column.labels=c("Pooled OLS", "Fixed Effects", "Fixed Effects", "Fixed Effects", "Random Effects"),
          add.lines=list(c("Time FE", "Yes", "Yes", "Yes", "Yes", "Yes"),
                         c("Individual FE", "No", "Yes", "Yes", "Yes", "No"),
                         c("Poverty and Gini controls", "No", "No", "No", "Yes", "No")),
          covariate.labels = c("Program dummy * 2000", "Program dummy * 2005", "Program dummy * 2015", "Program dummy * 2020", "Indigenous language speakers"),
          dep.var.labels="School attendance (6 to 11 years old)", no.space=TRUE, header=FALSE)
```


### 5.2.2. School enrollment

I iterate the same procedure using school enrollment as the outcome variable. The pooled OLS specification yields a 6.78 increase in school enrollment (significant at the 10% level). I find no statistically significant differences in placebo years, though they show relatively big coefficients and big standard errors.


```{r include=FALSE}
### Event Study:

temp_es <- f911

temp_es$time_to_treat <- temp_es$anio - 2014

modelo_tt <- feols(alum_tot ~ i(time_to_treat, trat, ref = -1) + factor(anio) + factor(mun), cluster = ~ mun, data = temp_es)
```


```{r results="asis", echo=FALSE, warning=FALSE}
iplot(modelo_tt, xlab = "Time to program", ylab = "ATT estimate (95% intervals)", main = "Figure 6. Event study of the program over school enrollment\nby school")
```


```{r include=FALSE}
temp <- f911

temp$dummy_00 <- 0
temp$dummy_00[temp$anio == 2000] <- 1
temp$dummy_01 <- 0
temp$dummy_01[temp$anio == 2001] <- 1
temp$dummy_02 <- 0
temp$dummy_02[temp$anio == 2002] <- 1
temp$dummy_03 <- 0
temp$dummy_03[temp$anio == 2003] <- 1
temp$dummy_04 <- 0
temp$dummy_04[temp$anio == 2004] <- 1
temp$dummy_05 <- 0
temp$dummy_05[temp$anio == 2005] <- 1
temp$dummy_06 <- 0
temp$dummy_06[temp$anio == 2006] <- 1
temp$dummy_07 <- 0
temp$dummy_07[temp$anio == 2007] <- 1
temp$dummy_08 <- 0
temp$dummy_08[temp$anio == 2008] <- 1
temp$dummy_09 <- 0
temp$dummy_09[temp$anio == 2009] <- 1
temp$dummy_10 <- 0
temp$dummy_10[temp$anio == 2010] <- 1
temp$dummy_11 <- 0
temp$dummy_11[temp$anio == 2011] <- 1
temp$dummy_12 <- 0
temp$dummy_12[temp$anio == 2012] <- 1
temp$dummy_14 <- 0
temp$dummy_14[temp$anio == 2014] <- 1
temp$dummy_15 <- 0
temp$dummy_15[temp$anio == 2015] <- 1
temp$dummy_16 <- 0
temp$dummy_16[temp$anio == 2016] <- 1
temp$dummy_17 <- 0
temp$dummy_17[temp$anio == 2017] <- 1
temp$dummy_18 <- 0
temp$dummy_18[temp$anio == 2018] <- 1
temp$dummy_19 <- 0
temp$dummy_19[temp$anio == 2019] <- 1

temp$diff_00 <- temp$trat * temp$dummy_00
temp$diff_01 <- temp$trat * temp$dummy_01
temp$diff_02 <- temp$trat * temp$dummy_02
temp$diff_03 <- temp$trat * temp$dummy_03
temp$diff_04 <- temp$trat * temp$dummy_04
temp$diff_05 <- temp$trat * temp$dummy_05
temp$diff_06 <- temp$trat * temp$dummy_06
temp$diff_07 <- temp$trat * temp$dummy_07
temp$diff_08 <- temp$trat * temp$dummy_08
temp$diff_09 <- temp$trat * temp$dummy_09
temp$diff_10 <- temp$trat * temp$dummy_10
temp$diff_11 <- temp$trat * temp$dummy_11
temp$diff_12 <- temp$trat * temp$dummy_12
temp$diff_14 <- temp$trat * temp$dummy_14
temp$diff_15 <- temp$trat * temp$dummy_15
temp$diff_16 <- temp$trat * temp$dummy_16
temp$diff_17 <- temp$trat * temp$dummy_17
temp$diff_18 <- temp$trat * temp$dummy_18
temp$diff_19 <- temp$trat * temp$dummy_19

temp$after_14 <- 0
temp$after_14[temp$anio >= 2014] <- 1
temp$diff <- temp$trat * temp$after_14

p_datos <- pdata.frame(temp, index = c("mun", "anio"))
is.pbalanced(p_datos)
p_datos <- p_datos[order(p_datos$mun, p_datos$anio),]
```


```{r include=FALSE}
### Matrícula total.

# 1. Pooled OLS: solo 2014.

#modelo1 <- lm(alum_tot ~ trat + diff + after_14, temp)
#cov1 <- vcovCR(modelo1, type="CR1", cluster=temp$mun)
#robust_se1 <- sqrt(diag(cov1))

# 2. Pooled OLS.

modelo2 <- lm(alum_tot ~ trat + diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), temp)
cov2 <- vcovCR(modelo2, type="CR1", cluster=temp$mun)
robust_se2 <- sqrt(diag(cov2))

# 3. Efectos Fijos.

p_datos <- pdata.frame(temp, index = c("mun", "anio"))
is.pbalanced(p_datos)
p_datos <- p_datos[order(p_datos$mun, p_datos$anio),]

modelo3 <- plm(alum_tot ~ diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), p_datos, model = "within")
cov3 <- vcovCR(modelo3, type="CR1", cluster=p_datos$mun)
robust_se3 <- sqrt(diag(cov3))

# 4. Efectos Aleatorios.

modelo4 <- plm(alum_tot ~ trat + diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), p_datos, model = "random")
cov4 <- vcovCR(modelo4, type="CR1", cluster=p_datos$mun)
robust_se4 <- sqrt(diag(cov4))

# Prueba de Hausman.

phtest(modelo3, modelo4)
```


```{r results="asis", echo=FALSE}
stargazer(modelo2, modelo3, modelo4,
          title="Differences in Differences over school enrollment",
          se=list(robust_se2, robust_se3, robust_se4),
          omit=c("trat", "anio", "Constant"),
          omit.stat=c("f", "adj.rsq", "ser"),
          font.size="tiny",
          column.sep.width="0pt",
          model.names=F,
          column.labels=c("Pooled OLS", "Fixed Effects", "Random Effects"),
          add.lines=list(c("Time FE", "Yes", "Yes", "Yes"),
                         c("Individual FE", "No", "Yes", "No")),
          covariate.labels = c("Program dummy * 2000", "Program dummy * 2001", "Program dummy * 2002", "Program dummy * 2003", "Program dummy * 2004", "Program dummy * 2005", "Program dummy * 2006", "Program dummy * 2007", "Program dummy * 2008", "Program dummy * 2009", "Program dummy * 2010", "Program dummy * 2011", "Program dummy * 2012", "Program dummy * 2014", "Program dummy * 2015", "Program dummy * 2016", "Program dummy * 2017", "Program dummy * 2018", "Program dummy * 2019"),
          dep.var.labels=c("School enrollment", "School enrollment", "School enrollment"), no.space=TRUE, header=FALSE)
```


\newpage


### 5.2.3. Number of teachers

Table 5 presents results for the number of teachers in indigenous education schools as the dependent variable. In this case, there seems to be a significant drop in the untreated group between 2017 and 2018, while the treated group remained fairly constant in the number of teachers.

Derived from this, I estimate a significant coefficient of 0.40, significant at the 5% level in 2018. However, for 2019, there seems to be a recovery by the control group, roughly equaling the average number of teachers in the program group.


```{r include=FALSE}
### Event Study:

modelo_tt <- feols(prof_tot ~ i(time_to_treat, trat, ref = -1) + factor(anio) + factor(mun), cluster = ~ mun, data = temp_es)
```


```{r results="asis", echo=FALSE, warning=FALSE}
iplot(modelo_tt, xlab = "Time to program", ylab = "ATT estimate (95% intervals)", main = "Figure 6. Event study of the program over number of teachers\nby school")
```


```{r include=FALSE}
### Número de docentes.

# 1. Pooled OLS: solo 2014.

#modelo1 <- lm(prof_tot ~ trat + diff + after_14, temp)
#cov1 <- vcovCR(modelo1, type="CR1", cluster=temp$mun)
#robust_se1 <- sqrt(diag(cov1))

# 2. Pooled OLS.

modelo2 <- lm(prof_tot ~ trat + diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), temp)
cov2 <- vcovCR(modelo2, type="CR1", cluster=temp$mun)
robust_se2 <- sqrt(diag(cov2))

# 3. Efectos Fijos.

p_datos <- pdata.frame(temp, index = c("mun", "anio"))
is.pbalanced(p_datos)
p_datos <- p_datos[order(p_datos$mun, p_datos$anio),]

modelo3 <- plm(prof_tot ~ diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), p_datos, model = "within")
cov3 <- vcovCR(modelo3, type="CR1", cluster=p_datos$mun)
robust_se3 <- sqrt(diag(cov3))

# 4. Efectos Aleatorios.

modelo4 <- plm(prof_tot ~ trat + diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), p_datos, model = "random")
cov4 <- vcovCR(modelo4, type="CR1", cluster=p_datos$mun)
robust_se4 <- sqrt(diag(cov4))

# Prueba de Hausman.

phtest(modelo3, modelo4)
```


```{r results="asis", echo=FALSE}
stargazer(modelo2, modelo3, modelo4,
          title="Differences in Differences over number of teachers",
          se=list(robust_se2, robust_se3, robust_se4),
          omit=c("trat", "anio", "Constant"),
          omit.stat=c("f", "adj.rsq", "ser"),
          font.size="tiny",
          column.sep.width="0pt",
          model.names=F,
          column.labels=c("Pooled OLS", "Fixed Effects", "Random Effects"),
          add.lines=list(c("Time FE", "Yes", "Yes", "Yes"),
                         c("Individual FE", "No", "Yes", "No")),
          covariate.labels = c("Program dummy * 2000", "Program dummy * 2001", "Program dummy * 2002", "Program dummy * 2003", "Program dummy * 2004", "Program dummy * 2005", "Program dummy * 2006", "Program dummy * 2007", "Program dummy * 2008", "Program dummy * 2009", "Program dummy * 2010", "Program dummy * 2011", "Program dummy * 2012", "Program dummy * 2014", "Program dummy * 2015", "Program dummy * 2016", "Program dummy * 2017", "Program dummy * 2018", "Program dummy * 2019"),
          dep.var.labels=c("Number of teachers", "Number of teachers", "Number of teachers"), no.space=TRUE, header=FALSE)
```


\newpage


### 5.2.4. Number of classrooms

Table 6 shows the results using the number of classrooms as an outcome. This variable proxies for school infrastructure. The presence of the program is associated with a statistically significant increase of 0.54 classrooms per school of indigenous education by 2019. The difference begins to become noticeable by 2015, and grows greater with the time. Prior to the application of the program, I consistently estimate small and non-statistically significant differences.


```{r include=FALSE}
### Event Study:

modelo_tt <- feols(aulas ~ i(time_to_treat, trat, ref = -1) + factor(anio) + factor(mun), cluster = ~ mun, data = temp_es)
```


```{r results="asis", echo=FALSE, warning=FALSE}
iplot(modelo_tt, xlab = "Time to program", ylab = "ATT estimate (95% intervals)", main = "Figure 6. Event study of the program over number of classrooms\nby school")
```


```{r include=FALSE}
### Número de aulas.

# 1. Pooled OLS: solo 2014.

#modelo1 <- lm(aulas ~ trat + diff + after_14, temp)
#cov1 <- vcovCR(modelo1, type="CR1", cluster=temp$mun)
#robust_se1 <- sqrt(diag(cov1))

# 2. Pooled OLS.

modelo2 <- lm(aulas ~ trat + diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), temp)
cov2 <- vcovCR(modelo2, type="CR1", cluster=temp$mun)
robust_se2 <- sqrt(diag(cov2))

# 3. Efectos Fijos.

p_datos <- pdata.frame(temp, index = c("mun", "anio"))
is.pbalanced(p_datos)
p_datos <- p_datos[order(p_datos$mun, p_datos$anio),]

modelo3 <- plm(aulas ~ diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), p_datos, model = "within")
cov3 <- vcovCR(modelo3, type="CR1", cluster=p_datos$mun)
robust_se3 <- sqrt(diag(cov3))

# 4. Efectos Aleatorios.

modelo4 <- plm(aulas ~ trat + diff_00 + diff_01 + diff_02 + diff_03 + diff_04 + diff_05 + diff_06 + diff_07 + diff_08 + diff_09 + diff_10 + diff_11 + diff_12 + diff_14 + diff_15 + diff_16 + diff_17 + diff_18 + diff_19 + factor(anio), p_datos, model = "random")
cov4 <- vcovCR(modelo4, type="CR1", cluster=p_datos$mun)
robust_se4 <- sqrt(diag(cov4))

# Prueba de Hausman.

phtest(modelo3, modelo4)
```


```{r results="asis", echo=FALSE}
stargazer(modelo2, modelo3, modelo4,
          title="Differences in Differences over number of classrooms",
          se=list(robust_se2, robust_se3, robust_se4),
          omit=c("trat", "anio", "Constant"),
          omit.stat=c("f", "adj.rsq", "ser"),
          font.size="tiny",
          column.sep.width="0pt",
          model.names=F,
          column.labels=c("Pooled OLS", "Fixed Effects", "Random Effects"),
          add.lines=list(c("Time FE", "Yes", "Yes", "Yes"),
                         c("Individual FE", "No", "Yes", "No")),
          covariate.labels = c("Program dummy * 2000", "Program dummy * 2001", "Program dummy * 2002", "Program dummy * 2003", "Program dummy * 2004", "Program dummy * 2005", "Program dummy * 2006", "Program dummy * 2007", "Program dummy * 2008", "Program dummy * 2009", "Program dummy * 2010", "Program dummy * 2011", "Program dummy * 2012", "Program dummy * 2014", "Program dummy * 2015", "Program dummy * 2016", "Program dummy * 2017", "Program dummy * 2018", "Program dummy * 2019"),
          dep.var.labels=c("Number of classrooms", "Number of classrooms", "Number of classrooms"), no.space=TRUE, header=FALSE)
```


\newpage


# 6. Conclusions

Literacy plays a fundamental role in human and social capital. Yet lack of access to literacy and education is disproportionately borne by Mexico's indigenous population. Between 2014 and 2021, a social program called Educational Inclusion and Equity Program provided subsidies to elementary schools of indigenous education. Using information at the municipal level, I estimate the impact of the program via a difference-in-differences framework.

I conclude that the presence of the program generated an increase of 3.42 percentage points in the average literacy of children aged 6 to 11 in the municipalities with indigenous education schools between 2010 and 2020. For each model specification, placebo effects between 2000 and 2010 were included in order to provide evidence on the parallel trends assumption. In general, the assumption seems plausible in light of the fact that placebo coefficients were not statistically significant.

I also find that the presence of the program is linked to an increase of 1.31 percentage points in school attendance per municipality and increases of 6.78 students in school enrollment and 0.54 classrooms per indigenous education school in program municipalities. This evidence is consistent with an improvement in school infrastructure as a cause for increased literacy.

Why are these results different from the evidence provided by De Hoyos, Estrada, and Vargas (2018) about general subsidies in public schools in Argentina? In part, because universal subsidies might not work, but targeting them at schools with significant needs could make a difference. Another reason could be that these subsidies may not have an impact over standardized test results, but rather over more "basic" learning such as literacy.

What this evidence indicates is that the Mexican educational system must be adapted to allow greater inclusion of the indigenous population in literacy processes, without leaving aside their cultural heritage. However, although this program was originally fragmented into four different programs in 2020, the component for indigenous education (now Program for Attention to the Diversity of Indigenous Education) disappeared in 2021. Education is a necessary condition for development, but it cannot be guaranteed without financial resources (especially where they are most needed).

At this point, a relevant question could arise: how much did the program cost or, rather, how cost-effective was this intervention? According to the CONEVAL Inventory of Social Programs, in 2020, the indigenous education component of the program spent \$86.89 million pesos (CONEVAL 2020). Considering the coverage of 817 municipalities, this represents an average of $106.35 thousand pesos per municipality. In terms of cost-effectiveness, each percentage point of literacy gained by municipality cost about \$31,096 pesos; an amount much lower than a GDP per capita.

Educational processes must be plural and adaptable to the context of each region. This not only represents a necessary condition for the development of historically discriminated sectors of the population, but it is also a minimum condition of social justice to guarantee equal access to social rights. Relatively simple interventions such as the Educational Equity and Inclusion Program can improve Spanish learning during the critical period (prior to 12 years old).


\newpage


# 7. Bibliography

Barrett, P., et al. (2019). *The Impact of School Infrastructure on Learning: A Synthesis of the Evidence*. Washington, DC: The World Bank.

Bruns, B. & Luque, J. (2015). Profesores Excelentes: Cómo mejorar el aprendizaje en América Latina y el Caribe. Washington, DC: The World Bank.

Bühmann, Dörthe & Trudell, Barbara. (2007). Mother tongue matters: local language as a key to effective learning. Francia: UNESCO.

CONEVAL. (2020). Inventario CONEVAL de Programas y Acciones Federales de Desarrollo Social. Obtained from https://www.coneval.org.mx/Evaluacion/IPFE/Paginas/historico.aspx

CONEVAL. (2020). Medición de la Pobreza: La Cohesión Social. Obtained from https://www.coneval.org.mx/Medicion/Paginas/Cohesion_Social.aspx

CONEVAL. (2020). Medición de la Pobreza: Pobreza a nivel municipio. Obtained from https://www.coneval.org.mx/Medicion/Paginas/Pobreza-municipal.aspx

De Hoyos, R. Estrada, R., and Vargas, M. (2018). *Predicting individual wellbeing through test scores: evidence from a national assessment in Mexico*. Policy Research Working Paper 8459, The World Bank.

De Hoyos, R., Holland, P., and Troiano, S. (2015). *Understanding the trends in learning outomes in Argentina, 2000 to 2012*. Policy Research Working Paper 7518, The World Bank.

DOF. (28 de febrero de 2019). Acuerdo número 04/02/19 por el que se emiten las Reglas de Operación del Programa para la Inclusión y la Equidad Educativa para el ejercicio fiscal 2019.

Duarte, Moreno, and Gargiulo. (2011). *Infraestructura Escolar y Aprendizajes en la Educación Básica Latinoamericana: Un análisis a partir del SERCE*. Banco Interamericano de Desarrollo.

Evans, M., et al. (2010). *Family scholarly culture and educational succed: books and schooling in 27 nations*. Research in Social Stratification and Mobility.

Ganuza, Natalia & Hedman, Christina. (2017). *The impact of mother tongue instruction on the development of biliteracy: evidence from Somali-Swedish bilinguals*. Applied Linguistics, I-25. Oxford University Press.

Hanushek, E., Piopiunik, M., and Wiederhold, S. (2014). *The value of smartet teachers: international evidence on teacher cognitive skills and student performance*. NBER Working Paper 20727.

INEGI. (2000). Censo General de Población y Vivienda 2000. Obtained from https://www.inegi.org.mx/programas/ccpv/2000/

INEGI. (2005). Conteo de Población y Vivienda 2005. Obtained from https://www.inegi.org.mx/programas/ccpv/2005/

INEGI. (2010). Censo de Población y Vivienda 2010. Obtained from https://www.inegi.org.mx/programas/ccpv/2010/

INEGI. (2015). Encuesta Intercensal 2015. Obtained from https://www.inegi.org.mx/programas/intercensal/2015/

INEGI. (2020). Censo de Población y Vivienda 2010. Obtained from https://www.inegi.org.mx/programas/ccpv/2020/

Miscevic-Kadijevic, Gordana. (2015). *TIMSS 2011: Relationship between self-confidence and cognitive achievement for Serbia and Slovenia*. Revista Electrónica de Investigación Educativa, 17(3), 109-115.

Narro, J., & Moctezuma, D. (2012). Analfabetismo en México: una deuda social. Obtenido de https://www.inegi.org.mx/rde/RDE_07/Doctos/RDE_07_Art1.pdf

Nkonde, et al. (2018). *Evaluating the impact of teaching and learning of mathematics and science using local language (language of play) in primary schools in Muchinga Province, Zambia, a case of Chinsali District*. American Journal of Educational Research, 6(8), pp. 1153-1163.

Reimers, Fernando. (2006). *Teaching quality matters. Pedagogy and literacy instruction of poor students in Mexico*. International Education for the Millennium. United States: Harvard Education Press.

Rivkin, S., Hanushek, E., and Kain, J. (2005). *Teachers, Schools and Academic Achievement*. Econometrica, Vol. 73, No. 2, pp. 417-458.

SEP. (2014). *Diagnóstico del Programa: Programa S244 Inclusión y Equidad Educativa*. Mexico: SEP.

Smits, Huisman, and Kruijff. (2008). *Home language and education in the developing world*. Netherlands: UNESCO.


